---
title: "R for DI consultants session 3"
author: "Steven Wink"
date: "8 juni 2016"
output: html_document
---


* With ```CTRL``` + ```ENTER``` you can execute commands from the .Rmd file in RStudio.

# Session 3: syntax & packages
* S3 vs S4
* tidy your data with 'tidyr'
* manipulate your data with 'dplyr'


## attributes and accessor functions
* attributes are metadata to objects/ structures
** names ```names()```
** dimensions ```dim()```
** class ```class()```

* To find out about the attributes use ```attributes()```

```{r, attributes and accessor functions}

a_vector <- c( 1, 2, 3, 3, 3 )
names( a_vector ) <- c("one", "two", "three", "three", "three")
a_vector
# or

also_a_vector <- c( one = 1, two = 2, three = 3, three = 3, three = 3 )
also_a_vector

identical( a_vector, also_a_vector )

a_vector <- unname( a_vector )

names( also_a_vector ) <- NULL

identical( a_vector, also_a_vector )


```

## factors as attribute
** used to store categorical data
** the ```class()``` is "factor"
** the ```levels()``` contains the set of allowed values

* categorical input data is usually required to be in the factor type for statistical functionality.
* can also be ordered for ordinal type of data

```{r, factors as attribute }

a_data_frame <- data.frame( a_vector = a_vector , also_a_vector = also_a_vector )
a_data_frame
#apply to each list entry ( column in the data.frame ), the class function:
lapply( a_data_frame, class )

a_data_frame <- lapply( a_data_frame, function( x ) x <- factor( x ) )

a_data_frame
lapply( a_data_frame, class )

# ordered ( ordinal data )

a_data_frame <- lapply( a_data_frame, function( x ) x <- factor( x, levels = c( 3, 2, 1), ordered = TRUE ) )

a_data_frame
levels( a_data_frame$a_vector )

```


## subsetting S4
** ```@``` is equivalent to ```$```
** ```slot()``` is equivalent to ```[[```

```{r subsetting S3 and S4 }





```


You can also embed plots, for example:

```{r, echo=FALSE}



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
