---
title: "R for DI consultants"
author: "Steven Wink"
date: "25 mei 2016"
output: html_document
---


The best resource for programmers to learn R is likely: http://adv-r.had.co.nz/
In the Subsetting and Vocabulary chapters you can quickly find most of the basic syntax and operations to start programming in R. 

The author 'Hadley Wickham' is a well known R developer, he spends most of his development work making R more accessible and syntax-standardized. He and his team have developed many data manipulation and graphics packages, and in this course we will have a look as some of these.

* With ```CTRL``` + ```ENTER``` you can execute commands from the .Rmd file in RStudio.

## Session 1: Basic operations
* functions, objects, some syntax
* generate data with specific distributions
* read and write
* data manipulation



This course is generated in a markdown file. A markdown file allows mixing up of formatted text and code 'chunks'.  

*Click the question mark above 'Quick Reference' to check some of the functionality.* Easy as ```1 + 1 = ``` `r 1 + 1`, right?

*Create a markdown file of your own*
*Set up the folders*


```{r, set up folders, warning = FALSE}

RootFolder <- "D:/leren/consultantR"  # change parent directory as you like

if( !dir.exists( RootFolder )) {
  dir.create(RootFolder)
}

setwd( RootFolder )

dir.create( "data" )
dir.create( "images" )
dir.create( "scripts" )
dir.create( "scripts/shiny" )

```

*Save your markdown file in the root folder*


## Introduction  

* R: strong foundation in both 'functional programming' and 'object-oriented' programming
* Programming language with metaprogramming facilities: programming + interactive data analysis functionality + write domain specific code/functions + connection to C, fortran etc.


* Main data storage types
* Function syntax


### vectors
* vectors contain homogenous data
* 1 dimensional

```{r, vectors}

1:10
c( 1:10 ) 
c( 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
c( 1, 2, 3, 4, 5, "6", 7, 8, 9, 10)

my_vector <- 1:10
my_vector / my_vector 
2 * my_vector
my_vector + my_vector

# beware of auto recycling

my_smaller_vector <- 1:5
my_vector + my_smaller_vector 


# only this will warn you:

my_3l_vector <- 1:3
my_vector + my_3l_vector

# subsetting vector:

my_vector[4]
my_vector[4:7]
my_vector[ -(3:6)]

```

### data.frame
* heterogenous data
* 2 dimensional
* most used data object
* is a list-type with some added functionality and some limitations (same length of lists)

```{r, data.frame}

a_dataframe = data.frame( firstColumn = 1:3,
                          secondColumn = c("row1", "row2", "row3"))
a_dataframe

# subsetting data.frame[ row, column ]
a_dataframe[ , 2 ] # all rows second column
a_dataframe[ , 2, drop = FALSE]  # # keep the original object type with drop = FALSE
a_dataframe[ , 'secondColumn']
a_dataframe[ , c( FALSE, TRUE )]  # vector based subsetting
a_dataframe[ , c( 1, 2 ) ]
a_dataframe$firstColumn
a_dataframe[ a_dataframe$firstColumn == 2,  ]
a_dataframe[ a_dataframe$firstColumn <= 2,  ]

# each column is a list, functions that work on lists work the same way on the columns of the data.frame:
# 'list apply'  apply function to each list entry
lapply( a_dataframe, unique)
lapply( a_dataframe, class )

# a dataframe with only numeric entries
a_numric_dataframe = data.frame( numCol1 = 1:10, numCol2 = sample(200:300, 10))
a_numric_dataframe
a_numric_dataframe + 1
a_numric_dataframe^2
lapply(a_numric_dataframe, function(x) x^2  ) # same as before.


```

### matrix
* for computational intensive calculations
* for linear algebra

```{r, matrix}

big_matrix = matrix( 1: 100000000, nrow = 1000000, ncol = 100 ) # 100M entries
format( object.size( big_matrix ), units = 'Mb') 
# inner product

big_matrix[ 960:989, 40:45 ] # subsetting ( $  and names does not work )

# math/ linear algebra
big_matrix2 <- big_matrix * 2

?'%*%'

big_matrix3 <- big_matrix / ( big_matrix2 ) + 1
dim( big_matrix3 )
big_matrix3[ 1:10, 1:10 ]

rm( "big_matrix", "big_matrix2", "big_matrix3" )

#inner product
1:10 %*% 1:10

#outer product
1:10 %o% 1:10

# product
1:10 * 1:10

```

### array
* multidimensional matrix

```{r, array}
some_data <- runif( 100*10*2)
my_array = array( data =  some_data, dim = c(100, 10, 2) )

my_array[ 1, 1, 1 ]
my_array[ 100, 1, 1 ]
my_array[ 1, 1,  ]
my_array[ 1, , 1  ]
my_array[ 1, ,   ]
my_array[ 88, ,   ]

```


### lists
* lists can contain heterogenous data. 
* lists can contain other data structures/ objects/ functions
* often used as input and/or output of functions because it can funciton as a container for various types of data


```{r, list}

my_list = alist( names = c( "John", "Steve" ) , age = as.vector( 22,44 ) )
my_list$names
my_list[1] # outputs object as subset of list
my_list[[1]] # outputs the object itself

my_list = alist() # define new empty list
my_list[[1]] <- a_dataframe
my_list[[2]] <- my_vector
myParameter <- 5
my_list[[3]] <- myParameter
another_list <- list( data = 1:10, name_data = "ten people")
my_list[[4]] <- another_list
my_list # That is why a list is usefull as input/output for functions

```


### generate data


```{r, generate data read & write }

# google: 'R generate names'  
#install.packages( 'randomNames' ) # install package, only needed once.
require( randomNames ) # load package, load every r-session you want to use it
??randomNames # how ? Aha! The ?? is a fuzzy search
# the single question mark is very specific and used most often.


# make a dataframe for african woman
?randomNames
set.seed(1234) # reproducible sampling of Genders:
Gender <- sample( c( "Male", "Female" ), 100, replace = TRUE )

my_dataframe_Woman <- data.frame( colAlphabet = sample( letters[ 1:3 ], 100, replace = TRUE ), 
                            Name = randomNames( n = 100, gender = Gender, ethnicity = 1), 
                            Ethnicity = "AfricanAmerican",
                            Gender = Gender,
                            age = round( 
                                rnorm( 
                                  100, mean = 40, sd = 10 ), digits = 0 ),
                            Length = round( runif( 100, min = 155, max = 189 ), digits = 2 )
)

head( my_dataframe_Woman )

write.table( my_dataframe_Woman, file = paste("data", "my_dataframe_Woman.txt", sep = "/"), sep ="\t", row.names = FALSE)

set.seed(543)
Gender <- sample( c( "Male", "Female" ), 100, replace = TRUE )
# make a dataframe for asian men
my_dataframe_Men <- data.frame( colAlphabet = sample( letters[ 1:3 ], 100, replace = TRUE ) , 
                            Name = randomNames( n = 100, ethnicity = 3),
                            Ethnicity = "Asian",
                            Gender = Gender,
                            age = round( 
                                rnorm( 
                                  100, mean = 40, sd = 10 ), digits = 0 ),
                            Length = round( runif( 100, min = 156, max = 204 ), digits = 2 )
)


head( my_dataframe_Men)
write.table( my_dataframe_Men, file = paste("data", "my_dataframe_Men.txt", sep = "/"), sep ="\t", row.names = FALSE)
# load the data in a list using a loop and combine into 1 dataframe

load.list = alist()
my.files <- paste( 'data', dir('data'), sep ='/' )

for( i in seq_along( my.files )) {
  
  load.list[[ i ]] <- read.delim( file= my.files[ i ] , sep = "\t" )
  
}

my_dataframe <- do.call( 'rbind', load.list ) # rbind the lists into a dataframe
my_dataframe[1:10,]

# we already used alot of built-in functions, and also the randomNames function from the loaded package:
?sample
?round
?rnorm
?randomNames # this function has the same name as the package
?head

# we also used some R objects:
?letters
?data.frame


# depending of the object type, secific methods on those objects are called, if implemented.

plot( my_vector )

plot(my_dataframe, col = as.factor( my_dataframe$Length > 200 ) )


```


### generate data distributions
* For test cases and or demo' s it can handy to be able to generate all kinds of data distributions
* base-R has several, such as use of sample(), ```runif()``` (random uniform), ```rnorm()``` (random normal), ```rbinom()``` (binomial), ```rpois()``` (poison)  
* also various packages exist that can generate specific (e.g. skewed) distributions of data  
* there are many example datasets available, in base-r and in packages, e.g. try ```mtcars```  


For any type of statistics (mean, variance, outlier detection, regression, etc) the data scientist must always be aware of the distribution

```runif``` equal probability of outcomes independant of the value  
for example throwing a die, the outcomes 1 through 6 have a runif distribution 
```{r, runif}
plot( density( runif( 1000, min = 1 , max = 6 ) ) ) 
hist( runif( 1000, min = 1 , max = 6 ) ) # likely a better method in this case

```

```rnorm``` naturally occuring distribution for many phenoma where the probability is highest around the mean and lower at the sides  
for example weight of a population
```{r, rnorm}
plot( density( rnorm( 1000, mean = 80 , sd = 10 ) ) )

```

```rbinom``` when there are two possible outcomes and one counts the number of occurences of the outcomes (chance of multiple succesive identical outcomes decreases)
```{r, rbinom}
plot( ( rbinom( 1000, size = 1, prob = 0.5 ) ) )
plot( density( rbinom( 1000, size = 1, prob = 0.5 ) ) )

plot( density( rbinom( 1000, size = 1, prob = 0.1 ) ) )

plot( density( rbinom( 10000, size = 6, prob = 0.3 ) ) ) # the chance for 6 succesive events with probability 0.3 is very small


```

```rpois``` chance of a fixed number of events in a given interval   
for example length of a line of people, or how long people have to wait certain time amounts at a bus station when they arive randomly (don' t check 9292ov.nl)  
```{r, rpois}

plot(density ( rpois( 1000, lambda = 5 ),bw = 0.1 )  ) # you expect 5 occurrences in your defined interval. The poison assumption tells you the occurrences of surrounding occurrences

```

### Example datasets   
* Instead of generating your own data, there are many datasets available within R and R-packages

```{r }

data() # available/installed datasets

mtcars
?mtcars
summary( mtcars )
plot( mtcars )
boxplot( mtcars)
boxplot( scale ( mtcars ) )

```



### R is built around functions
* scoping: within functions is searched first, if none is found global variables are searched
* Open a new R Script (File -> New File -> R Script)
* Write the following function in it and save as weeksOld.R in your working directory

```{r, r-script function}
# Write the following in the .R file:


weeksOld <- function( Birth_date ) {
  Nweeks = alist()  
    Nweeks$WeeksAge <- difftime( Sys.Date(), as.Date( Birth_date, format = "%Y-%m-%d" ), units = "weeks"  )
    Nweeks$Today <- Sys.Date()
    Nweeks$Born <- Birth_date
    Nweeks$call <- call("as.Date", as.Date(Birth_date, format = "%Y-%m-%d" ) )
return( Nweeks )
}



```

### test our function:

```{r , functions}

source("weeksOld.R")

function_result <- weeksOld( "1980-04-02" )

function_result

# scoping rules of functions
my_var <- 10

my_fun <- function( x ) {
  y <- x + my_var
  return( y )
}
my_fun( 1 )


my_fun <- function( x ) {
   my_var = 0
  y <- x + my_var
  return( y )
}
my_fun( 1 )


my_fun <- function( x, my_var ) {
    y <-   x + my_var
    my_var = 0
   return( y )
  }



my_fun( 1, 2 )


```

### data manipulation
* base-r has several data managing methods  
* there are however often better/ easier/ faster/ more concise methods to be found in various packages ( one of the next sessions )

```{r, data manipulation}

# summarize data with aggregate

head(my_dataframe)

aggregate( age ~ Ethnicity + Gender , FUN = mean, data = my_dataframe)

my_dataframe_stack <- stack( x = my_dataframe, select = c( age, Length ) )
head( my_dataframe_stack )
# data in long format with base-r function reshape. Difficult to grasp but very flexible
# easier/ prefered methods from packages in one of next sessions.
my_dataframe_long <- reshape( my_dataframe,
         direction = "long",
         varying =c("age", "Length", "Gender", "Ethnicity", "colAlphabet"),
         v.names = "value",
         times =c("age", "Length", "Gender", "Ethnicity", "colAlphabet"),
         timevar = "Info")

head( my_dataframe_long )


```


### Outlier detection

```{r, preview outlier detection }
# dataset on flow of river Nile
# statistical method to detect outliers?

plot(Nile)
hist(Nile)
# looks normal distributed so we could try 2*SD?

SD_Nile <- 2* sd(Nile)

Outliers1 <- Nile[ Nile > ( mean( Nile ) + SD_Nile ) | Nile < ( mean( Nile ) - SD_Nile ) ]  # 
Nile
plot(Nile)

points(x =  ( 1871:1970)[ Nile %in% Outliers1] , y = Outliers1 , cex =5, col = "red")

# But there is a global trend blowing up the SD at certain locations and underestimating at other locations.
# Better would be to calculate the values with respect to this trend:

Nile.df <- data.frame( Nile = as.numeric( Nile ), Time = 1871:1970)

lin.fit <- lm( Nile ~ Time + I( Time^2 ), data = Nile.df   ) # fit 2nd degree polynomial model (regression)

lin.fit.s <- summary( lin.fit )

Nile.df$fittedValue <- lin.fit.s$coefficient[ 1 ] + lin.fit.s$coefficient[ 2 ] * Nile.df$Time + lin.fit.s$coefficient[ 3 ] * Nile.df$Time^2

plot( x = Nile.df$Time, y = Nile.df$Nile, pch = 1, col = "blue" )

lines( Nile.df$Time, Nile.df$fittedValue, col = "red" )

# or with the ```predict``` function:

points( x = 1871:1970, y =predict( lin.fit )  , col = "lightblue", cex = 0.5, pch = 'X')

futureTime = data.frame(Time = 1871:2200)

# a 'very accurate' prediction for the year 2200

plot( x = Nile.df$Time, y = Nile.df$Nile, pch = 1, col = "blue" , xlim = c(1871, 2200 ), ylim = c(0,6000))
points(  x = futureTime$Time, y = predict( lin.fit , newdata = futureTime ), col = 'darkgreen' )

# the residual standard error
resi.sd <- sqrt( sum( lin.fit.s$residuals^2 ) / ( 98- 1)) 

Outliers2 <- Nile.df$Nile[ abs( lin.fit.s$residuals ) > 2 * resi.sd]

Outliers2

plot(Nile)
lines( Nile.df$Time, Nile.df$fittedValue, col = "red" )
points(x =  ( 1871:1970)[ Nile %in% Outliers2] , y = Outliers2 , cex =5, col = "red")


```





### Exercies session 1


1. calculate the maximum length for each age using the ```aggregate``` function  
2. what is the maximum number of occurences of the age group? ( use ```FUN = length``` )  
3. write a function in a R-script with:
*** input1: a ```rnorm``` generated vector of length 100 with mean 3 and sd 2
*** input2: a ```rnorm``` generated vector of length 100 with mean 7 and sd 3
*** in the function: put input2 behind input1, plot and perform a regression (pro tip: r functions like data.frames)
*** output: a plot in png format with the rnorm generated data on the y-axis (and 1:200 on the x axis)
*** the plot should contain a regression line lm(a~x, data = ...) is linear lm(a~ I(x + x^2 + x^3 + x^4), data = ...) is non a non linear regression )
tip: ```poly( x, degree = 4 )``` makes the polynomial for you: ``` a ~ poly(x, degree = ...)```
4. what happens when you increase the degrees of freedom? ( by increasing the polynomial degree )





